---
title: "FRE 6871 - Final Project: Predicting next-day SPY returns and direction"
author: "Tong Mo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: readable
    highlight: tango
    df_print: paged
    code_folding: show
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo    = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      fig.width = 9, fig.height = 6,
                      cache    = FALSE)
```

# Introduction & research question

Equity ETFs are the most liquid expression of broad-market views. This
project asks two related questions about **SPY** (the S&P 500 ETF):

1. **Magnitude.** Can we forecast the *size* of next-day log returns
   from a small set of features built from SPY itself, four cross-asset
   ETFs (QQQ, IWM, GLD, TLT) and two macro indicators (^VIX, ^TNX)?
2. **Direction.** Can we beat a 50/50 coin-flip on next-day *direction*
   using logistic regression and SVM, and does either of them
   outperform a pure ARIMA forecast?

References to the textbook (Kabacoff, *R in Action*, 3rd edition) are
flagged inline at the first use of each technique.

```{r packages}
suppressPackageStartupMessages({
    library(quantmod)
    library(xts)
    library(zoo)
    library(PerformanceAnalytics)
    library(psych)
    library(car)
    library(MASS)
    library(boot)
    library(forecast)
    library(tseries)
    library(NbClust)
    library(flexclust)
    library(e1071)
    library(pROC)
    library(corrgram)
})
```

# Data acquisition

We use `quantmod::getSymbols()` (R in Action, ch. 2 - importing data)
to pull adjusted-close prices from Yahoo Finance.  Once successful, we
cache the merged data set to a CSV so subsequent runs do not need
internet.

```{r data-acquisition}
csv_path <- "Tong_Mo_Project_data.csv"
tickers  <- c("SPY", "QQQ", "IWM", "GLD", "TLT", "^VIX", "^TNX")

load_data_online <- function() {
    env <- new.env()
    getSymbols(tickers, from = "2010-01-01", to = "2024-12-31",
               env = env, auto.assign = TRUE)
    cols <- list(SPY = Ad(env$SPY), QQQ = Ad(env$QQQ), IWM = Ad(env$IWM),
                 GLD = Ad(env$GLD), TLT = Ad(env$TLT),
                 VIX = Cl(env$VIX), TNX = Cl(env$TNX))
    px <- do.call(merge, cols)
    colnames(px) <- names(cols)
    px
}

if (file.exists(csv_path)) {
    cached <- read.csv(csv_path, stringsAsFactors = FALSE)
    px <- xts(cached[, -1], order.by = as.Date(cached[, 1]))
} else {
    px <- load_data_online()
    write.csv(data.frame(Date = index(px), coredata(px)),
              file = csv_path, row.names = FALSE)
}

cat("Series span:", format(start(px)), "to", format(end(px)), "\n")
cat("Trading days:", nrow(px), "\n")
```

# Initial exploration

```{r explore}
str(px)
head(px, 3)
summary(px)
```

```{r explore-plots}
plot.zoo(px[, c("SPY","QQQ","IWM","GLD","TLT")],
         main = "Adjusted prices - five ETFs (2010-2024)",
         xlab = "Date", col = c("steelblue","darkorange","forestgreen",
                                 "goldenrod","firebrick"))
plot.zoo(px[, c("VIX","TNX")],
         main = "Macro indicators - VIX and 10Y yield",
         xlab = "Date", col = c("firebrick","steelblue"))
```

# Cleaning & missing values

R in Action sec. 4.5 / 15.1 covers handling missing data. VIX and TNX
have a few non-trading-day NAs; we carry the last value forward and
drop any row that still has a missing value, logging the impact.

```{r clean}
n_before <- nrow(px)
colSums(is.na(px))
px       <- na.locf(px, na.rm = FALSE)
px       <- px[complete.cases(px), ]
n_after  <- nrow(px)
sprintf("Rows dropped during cleaning: %d (%.2f%%)",
        n_before - n_after, 100 * (n_before - n_after) / n_before)
colSums(is.na(px))
```

# Feature engineering

```{r features}
ret <- na.omit(Return.calculate(px, method = "log"))
colnames(ret) <- paste0(colnames(ret), "_ret")

vol20 <- runSD(ret$SPY_ret, n = 20) * sqrt(252)
colnames(vol20) <- "vol20"

lag_features <- merge(
    lag(ret$SPY_ret, 1), lag(ret$SPY_ret, 2), lag(ret$SPY_ret, 3),
    lag(ret$QQQ_ret, 1), lag(ret$IWM_ret, 1),
    lag(ret$GLD_ret, 1), lag(ret$TLT_ret, 1),
    lag(ret$VIX_ret, 1), lag(ret$TNX_ret, 1),
    lag(vol20, 1))
colnames(lag_features) <- c("spy_lag1","spy_lag2","spy_lag3",
                            "qqq_lag1","iwm_lag1","gld_lag1","tlt_lag1",
                            "vix_lag1","tnx_lag1","vol20_lag1")

y       <- ret$SPY_ret
y_dir   <- xts(as.integer(coredata(y) > 0), order.by = index(y))
colnames(y) <- "spy_ret"
colnames(y_dir) <- "spy_up"

dataset <- na.omit(merge(y, y_dir, lag_features))
df      <- data.frame(date = index(dataset), coredata(dataset))
df$spy_up <- factor(df$spy_up, levels = c(0, 1), labels = c("dn", "up"))

# 80/20 chronological split
split <- floor(0.8 * nrow(df))
train <- df[1:split, ]
valid <- df[(split + 1):nrow(df), ]
sprintf("train: %d (%s -> %s)  | valid: %d (%s -> %s)",
        nrow(train), train$date[1], train$date[nrow(train)],
        nrow(valid), valid$date[1], valid$date[nrow(valid)])
```

# Descriptive statistics & basic plots

R in Action ch. 7 - `psych::describe` (7.1.1) and `corrgram` (11.3.1).

```{r descriptive}
psych::describe(coredata(ret))[, c("mean","sd","skew","kurtosis")]

hist(coredata(ret$SPY_ret), breaks = 60, freq = FALSE,
     col = "steelblue", border = "white",
     main = "Distribution of SPY daily log returns", xlab = "log return")
lines(density(coredata(ret$SPY_ret)), col = "firebrick", lwd = 2)

corrgram(coredata(ret), order = TRUE,
         lower.panel = corrgram::panel.shade,
         upper.panel = corrgram::panel.pie,
         text.panel  = corrgram::panel.txt,
         main = "Cross-asset return correlations")

chart.CumReturns(ret$SPY_ret, main = "SPY cumulative log return")
```

# Statistical tests

We run **five** formal tests; only two are required by the rubric.

## Normality (R in Action 7.5)

```{r test-normality}
spy_vec <- as.numeric(coredata(ret$SPY_ret))
shapiro.test(if (length(spy_vec) > 5000) sample(spy_vec, 5000) else spy_vec)
```

Returns are heavy-tailed (p \<\< 0.05 - reject normality).

## Stationarity - prices vs returns (R in Action 15.2.1)

```{r test-adf}
suppressWarnings(adf.test(coredata(px$SPY)))
suppressWarnings(adf.test(coredata(ret$SPY_ret)))
```

Prices fail to reject the unit-root null; returns clearly reject it -
ARIMA must be applied to returns.

## Mean return = 0?  (R in Action 7.4)

```{r test-tt}
t.test(coredata(ret$SPY_ret))
wilcox.test(coredata(ret$SPY_ret))
```

Both tests reject zero mean - the unconditional drift is positive.

## Day-of-week vs direction (R in Action 7.3)

```{r test-chisq}
wday <- weekdays(index(ret$SPY_ret))
tab  <- table(weekday = wday,
              direction = ifelse(coredata(ret$SPY_ret) > 0, "up", "dn"))
tab
chisq.test(tab)
```

## Box-Ljung autocorrelation

```{r test-box}
Box.test(coredata(ret$SPY_ret), lag = 10, type = "Ljung-Box")
```

# Modelling

## Multiple linear regression - R in Action ch. 8

```{r model-lm}
fit_lm_full <- lm(spy_ret ~ spy_lag1 + spy_lag2 + spy_lag3 +
                              qqq_lag1 + iwm_lag1 + gld_lag1 + tlt_lag1 +
                              vix_lag1 + tnx_lag1 + vol20_lag1,
                  data = train)
summary(fit_lm_full)

fit_lm_step <- stepAIC(fit_lm_full, direction = "both", trace = FALSE)
formula(fit_lm_step)
summary(fit_lm_step)

par(mfrow = c(2, 2)); plot(fit_lm_step); par(mfrow = c(1, 1))
ncvTest(fit_lm_step)

lm_pred  <- predict(fit_lm_step, newdata = valid)
lm_rmse  <- sqrt(mean((valid$spy_ret - lm_pred)^2))
lm_dirAcc <- mean(sign(lm_pred) == sign(valid$spy_ret))
sprintf("[lm] validation RMSE = %.6f, direction accuracy = %.3f",
        lm_rmse, lm_dirAcc)
```

## Bootstrap of the Sharpe ratio - R in Action ch. 12

```{r model-boot}
sharpe_stat <- function(data, idx) {
    r <- data[idx]
    sqrt(252) * mean(r) / sd(r)
}
set.seed(1234)
boot_out <- boot(coredata(ret$SPY_ret), statistic = sharpe_stat, R = 1000)
boot_out
boot.ci(boot_out, type = c("perc", "bca"))
```

## Logistic regression - R in Action sec. 13.2 *(NEW technique)*

```{r model-glm}
fit_glm <- glm(spy_up ~ spy_lag1 + spy_lag2 + spy_lag3 +
                          qqq_lag1 + iwm_lag1 + gld_lag1 + tlt_lag1 +
                          vix_lag1 + tnx_lag1 + vol20_lag1,
               data = train, family = binomial)
summary(fit_glm)

glm_prob <- predict(fit_glm, newdata = valid, type = "response")
glm_pred <- factor(ifelse(glm_prob > 0.5, "up", "dn"),
                   levels = c("dn", "up"))
glm_cm   <- table(actual = valid$spy_up, predicted = glm_pred)
glm_cm
glm_acc  <- sum(diag(glm_cm)) / sum(glm_cm)
glm_auc  <- as.numeric(pROC::auc(pROC::roc(valid$spy_up, glm_prob,
                                           quiet = TRUE)))
sprintf("[glm] validation accuracy = %.3f, AUC = %.3f", glm_acc, glm_auc)
```

## ARIMA - R in Action sec. 15.4

```{r model-arima}
spy_train_xts <- xts(train$spy_ret, order.by = train$date)
fit_arima <- auto.arima(spy_train_xts)
fit_arima
accuracy(fit_arima)

arima_fc   <- forecast(fit_arima, h = nrow(valid))
arima_pred <- as.numeric(arima_fc$mean)
arima_rmse <- sqrt(mean((valid$spy_ret - arima_pred)^2))
arima_dirAcc <- mean(sign(arima_pred) == sign(valid$spy_ret))
sprintf("[arima] validation RMSE = %.6f, direction accuracy = %.3f",
        arima_rmse, arima_dirAcc)
plot(forecast(fit_arima, h = 60),
     main = "ARIMA forecast - 60 trading days ahead")
```

## K-means clustering of weekly volatility regimes - R in Action ch. 16

```{r model-kmeans}
weekly <- apply.weekly(ret$SPY_ret,
                       function(x) c(mean = mean(x), sd = sd(x)))
weekly        <- na.omit(weekly)
weekly_scaled <- scale(coredata(weekly))

set.seed(1234)
nc     <- NbClust(weekly_scaled, min.nc = 2, max.nc = 6, method = "kmeans")
best_k <- as.integer(names(sort(table(nc$Best.n[1, ]), decreasing = TRUE))[1])
if (is.na(best_k) || best_k < 2) best_k <- 3
sprintf("NbClust majority vote: k = %d", best_k)

set.seed(1234)
fit_km <- kmeans(weekly_scaled, centers = best_k, nstart = 25)
fit_km$size
round(fit_km$centers, 3)

regime <- data.frame(date  = index(weekly),
                     mean  = coredata(weekly)[, "mean"],
                     sd    = coredata(weekly)[, "sd"],
                     regime= factor(fit_km$cluster))
aggregate(cbind(mean, sd) ~ regime, data = regime, FUN = mean)

plot(regime$sd, regime$mean, col = regime$regime, pch = 19,
     xlab = "weekly std-dev", ylab = "weekly mean return",
     main = paste("SPY weekly volatility regimes (k =", best_k, ")"))
legend("topright", legend = paste("regime", seq_len(best_k)),
       col = seq_len(best_k), pch = 19)
```

## SVM classification - R in Action sec. 17.5

```{r model-svm}
set.seed(1234)
fit_svm <- svm(spy_up ~ spy_lag1 + spy_lag2 + spy_lag3 +
                          qqq_lag1 + iwm_lag1 + gld_lag1 + tlt_lag1 +
                          vix_lag1 + tnx_lag1 + vol20_lag1,
               data = train, probability = TRUE)
summary(fit_svm)

svm_pred <- predict(fit_svm, newdata = valid, probability = TRUE)
svm_prob <- attr(svm_pred, "probabilities")[, "up"]
svm_cm   <- table(actual = valid$spy_up, predicted = svm_pred)
svm_cm
svm_acc  <- sum(diag(svm_cm)) / sum(svm_cm)
svm_auc  <- as.numeric(pROC::auc(pROC::roc(valid$spy_up, svm_prob,
                                           quiet = TRUE)))
sprintf("[svm] validation accuracy = %.3f, AUC = %.3f", svm_acc, svm_auc)

roc_glm <- pROC::roc(valid$spy_up, glm_prob, quiet = TRUE)
roc_svm <- pROC::roc(valid$spy_up, svm_prob, quiet = TRUE)
plot(roc_glm, col = "steelblue", lwd = 2,
     main = "ROC - logistic regression vs SVM")
lines(roc_svm, col = "firebrick", lwd = 2)
legend("bottomright", c(sprintf("logit  AUC=%.3f", glm_auc),
                          sprintf("svm    AUC=%.3f", svm_auc)),
       col = c("steelblue","firebrick"), lwd = 2)
```

# Model comparison

```{r comparison}
data.frame(
    model     = c("OLS (stepwise)", "ARIMA", "Logistic", "SVM"),
    target    = c("return",  "return",     "direction",  "direction"),
    metric    = c("RMSE",     "RMSE",      "Accuracy",   "Accuracy"),
    value     = c(lm_rmse,    arima_rmse,  glm_acc,      svm_acc),
    secondary = c(lm_dirAcc,  arima_dirAcc, glm_auc,     svm_auc),
    secondary_label = c("dir.Acc","dir.Acc","AUC","AUC"))
```

# Conclusion

1. **Magnitude prediction is hard.** OLS (RMSE = `r round(lm_rmse, 5)`)
   and ARIMA (RMSE = `r round(arima_rmse, 5)`) both produce a daily
   forecast error roughly the size of one ordinary daily move.
2. **Direction prediction is only marginally tractable.** Logistic
   reaches accuracy = `r round(glm_acc, 3)` / AUC =
   `r round(glm_auc, 3)`; SVM reaches `r round(svm_acc, 3)` /
   `r round(svm_auc, 3)`. Both sit close to the 50% coin-flip.
3. **Volatility regimes are real.** NbClust selects k =
   `r best_k`; the high-vol cluster has *negative* mean weekly return
   while the low-vol cluster has positive mean - a 'panic vs calm'
   dichotomy that is well documented in the literature.
4. **Statistical tests support the modelling choices** - non-normal
   returns, prices non-stationary but returns stationary, positive
   drift confirmed by both parametric and non-parametric tests.
5. **Future work** - technical indicators (RSI, MACD), a credit-spread
   feature, longer look-backs, gradient-boosted trees, and walk-forward
   evaluation.

------------------------------------------------------------------------

> *The .R version of this report (`Tong_Mo_Project.R`) is the primary
> deliverable; this `.Rmd` / HTML report is the optional dynamic-
> report version described in the FRE 6871 walkthrough.*
